\MakeUppercase{The Structure of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-07-03 k-planar.tex
Moving....

\renewcommand{\proplabel}[1]{\label{prop:#1}}
\renewcommand{\propref}[1]{(PR\ref{prop:#1})}
\newcommand{\jlabel}[1]{\label{j:#1}}
\newcommand{\jref}[1]{(J\ref{j:#1})}

@@

>>>>

\renewcommand{\proplabel}[1]{\label{prop:#1}}
\renewcommand{\propref}[1]{(PR\ref{prop:#1})}
\newcommand{\jlabel}[1]{\label{j:#1}}
\newcommand{\jref}[1]{(J\ref{j:#1})}
\newcommand{\tref}[1]{(T\ref{#1})}
\newcommand{\ylabel}[1]{\label{y:#1}}
\newcommand{\yref}[1]{(Y\ref{y:#1})}

@@

\newpage
\section{A Generalization}

We now generalize \thmref{k-planar} to a more abstract setting.  For two sets $A$ and $B$, we use $A\ominus B := (A\cup B)\setminus(A\cap B)$ to denote the symmetric difference between $A$ and $B$.  We begin with a lemma about normalized tree decompositions:

\begin{lem}\lemlabel{nice-decomposition}
  Let $H$ be a graph of treewidth $t-1$.  Then there exists a graph $H'$ isomorphic to $H$ that has a tree-decomposition $(B_x:x\in V(T))$ of width at most $t-1$ that has following additional properties:  
  \begin{compactenum}[(T1)]
    \item\label{t:rooted}\label{t:first} $T$ is rooted at some node $x_0\in V(T)$ with $B_{x_0}=\emptyset$.
    \item\label{t:node-set} $V(H')\subseteq V(T)$.
    \item\label{t:diff} For each edge $xy\in E(T)$, $|B_x\ominus B_y||\le 1$.
    \item\label{t:subtree-root} For each $v'\in V(H')$, the subtree $T[v']:=T[\{x\in V(T):v'\in B_x\}]$ is rooted at $v'$.
    \item\label{t:ancestor-edge}\label{t:last} For each edge $v'w'\in E(H)$, one of $v'$ or $w'$ is a $T$-ancestor of the other.
  \end{compactenum}
\end{lem}

\begin{proof}
  Begin with a width-$(t-1)$ tree-decomposition $(C_x:x\in V(T_0))$ of $H$ using a tree $T_0$ with the property that, for any edge $xy\in E(T_0)$, $|C_x\ominus C_y|\le 1$. That is, $C_x$ contains exactly one vertex not in $C_y$ or vice-versa. It is well-known that such a tree-decomposition exists.

  Next, select an arbitrary node $x\in V(T_0)$ and let $C_x:=\{v_1,\ldots,v_{t'}\}$ denote the contents of $C_x$.
  Create a new tree $T$ from $T_0$ by adding a path of nodes $x_0,x_1,x_2,\ldots,x_{t'-1},x_{t'}=x$. Let $C_{x_0}:=\emptyset$ and, for each $i\in\{1,\ldots,t'\}$, $C_{x_i}:=\{v_1,\ldots,v_i\}$.  Note that $(C_x: x\in V(T))$ is a width-$(t-1)$ tree-decomposition of $H$. Let $T$ be rooted at $x_0$. Note that $(C_x:x\in V(T))$ satisfies Properties~\tref{t:rooted}, and \tref{t:diff}.  For the remaining properties some renaming is required, which we now describe.

  For each $v\in V(H)$, let $v'\in V(T)$ be the unique node such that $v\in C_{v'}$ and $v'$ has minimum $T$-depth.  The uniqueness of $v'$ follows from the fact that $T[v]:=T[\{x\in V(T):v\in B_x\}]$ is connected, so $v'$ is the root of $T[v]$.  Furthermore, for any pair of distinct vertices $v,w\in V(H)$, \tref{t:rooted} and \tref{t:diff} ensure that $v'\neq w'$ since, otherwise, $C_{v'}\ominus C_{x'}\supseteq \{v',w'\}$ where $x'$ is the parent of $v'=w'$.  Property~\tref{t:rooted} would ensure that $x'$ exists and Property~\tref{t:diff} would ensure that $|C_{x'}\ominus C_{v'}|\le 1$.

  For each $x\in V(T)$, let $B_x=\{v' : v\in C_x\}$.  Now define the graph $H'$ to have vertex set $V(H')=\{v':v\in V(H)\}$ and edge set $E(H')=\{v'w': vw\in E(H)\}$.  Obviously, $H'$ isomorphic to $H$ and $V(H')\subseteq V(T)$ and $(B_x: x\in V(T))$ is a width=$(t-1)$ tree-decomposition of $H'$.  The decomposition $(B_x: x\in V(T))$ inherits Properties~\tref{t:rooted} and \tref{t:diff} from $(C_x:x\in V(T))$ and satisfies Property~\tref{t:node-set} and \tref{t:subtree-root} by definition.

  To see that $(B_:x\in V(T))$ satisifies Property~\tref{t:ancestor-edge}, observe that, if $vw\in E(H)$, then at least one of $v'$ or $w'$ is contained in $B_x$ for every node $x$ on the path from $v'$ to $w'$ in $T$.  If neither $v'$ nor $w'$ is an ancestor of the other, then some node $x$ on this path has $T$-depth less than that of $v'$ and $w'$.  If $v'\in B_x$ this contradicts the fact that $v'$ is the root of $T[v']$.  If $w'\in B_x$ this contradicts the fact that $w'$ is the root of $T[w']$.
\end{proof}

\newpage
\subsection{Generalized Tripod Decompositions}

Let $G$ be a graph and let $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{P}:=(Q_x: x\in V(H)))$ be a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{P}$ is at most $t$.  In this section we define a hierarchical structure on $G$ that can be obtained from $(\mathcal{L},\mathcal{P})$.

\begin{thm}
  There exists a tree $T$ rooted at a node $x_0$ where each node $x\in V(T)$ is associated with a subset $Y_x\subseteq V(G)$ that defines several related sets $V_x = \bigcup_{y\in V(T_x) Y_y}$, $F_x=\{w\in V(G): vw\in E(G), v\in V_x,\, w\not\in V_x\}$, and $N_x=V_x\cup F_x$ that satisfy the following properties:
  \begin{compactenum}[(Y1)]
    \item\ylabel{thickness} $\mathcal{Y}:=(Y_x: x\in V(T))$ is a partition of $V(G)$ of layered width (with respect to $\mathcal{L}$) at most $t\ell$.
    \item\ylabel{separator} There is no edge $vw\in E(G)$ with $v\in V_x$ and $w\in V(G)\setminus N_x$. 
    \item\ylabel{ancestor-edge} $x$ has $t'\le t$ $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{compactenum}
\end{thm}

\begin{proof}
  Apply \lemref{nice-decomposition} to $I$ to obtain a tree-decomposition $(B_x: x\in V(T))$ of $I$ where $T$ is rooted at some node $x_0$ and that satisifies Properties~\tref{t:first}--\tref{t:last}. (We skip the introduction of the isomorphic graph $H'$ and assume from the outset that $V(I)\subseteq V(T))$.)

  For each $v\in V(G)$, let $q(v)$ denote the unique node $q\in V(I)$ such that $v\in Q_q$ and let $b(v)$ denote the root of the subtree $T[v]:=T[\{x\in V(T):q(v)\in B_x]$. For each $x\in V(T)$, define $Y_x=\{v\in V(G):c(v)=x\}$.  
  Clearly, $\mathcal{Y}:=(Y_x:x\in V(T))$ is a partition of $V(G)$.  Furthermore, for any $i\in\in\{1,\ldots,h\}$ 
  \[  |Y_x| = \sum_{y\in C_x}|Q_y\cap V_i|\le |C_x|\ell \le t\ell ,
  \] 
  so $\mathcal{Y}$ has layered width at most $t\ell$ with respect to $\mathcal{L}$.  This establishes \yref{thickness}.

  Property \yref{separator} is immediate from the definitions of $F_x$ and $N_x$.  In particular, $(N_x,V(G)\setminus V_x)$ is a separation of $G$ with $F_x=N_x\cap(V(G)\setminus V_x)$.

  To establish Property~\yref{ancestor-edge}, consider some edge $vw\in E(G)$ with $v\in V_x$ and $w\not\in V_x$.  By the definition of $V_x$, $v\in Y_{x'}$ for some $T$-descendant $x'$ of $x$ (possibly $x=x'$). 
  Since $\mathcal{Y}$ is a partition, $w\in Y_{a}$ for some $a\in V(T)$, $a\neq x$.  Since $v\in Y_{x'}$ and $w\in Y_a$, $q(v)=x'$ and $q(w)=a$, so $x'\in Q_v$ and $w\in Q_a$.  Therefore $H$ contains the edge $ax'$.

  By \tref{ancestor-edge}, $a$ is an ancestor of $x'$  
\end{proof}

\newpage
\subsection{Low-Degree Powers}

For non-negative integers $k$ and $d$, a graph $G$ and distinct vertices $v,w\in V(G)$, a \emph{$(k,d)$ $vw$-shortcut} is a path in $G$ with endpoints $v$ and $w$ of length at most $k$ and whose internal vertices each have degree at most $d$.  The \emph{$(k,d)$th power of $G$}, denoted $G^{k,d}$, is the graph with vertex $V(G^{k,d})=V(G)$ in which, for each distinct pair $v,w\in V(G)$, $vw\in E(G^{d,k})$ if and only if $G$ contains a $(k,d)$ $vw$-shortcut.

\begin{thm}
  Let $\ell\ge 1$, $t\ge 2$, $d\ge 1$, and $k\ge 1$ be integers, let $G$ be a graph having a layered $I$-partition of layered width $\ell$ in which $I$ has treewidth at most $t-1$.  Then $G^{k,d}$ has a layered $H$-partition of layered width $O(\ell kd^k)$ where the treewidth of $H$ is at most $\binom{t+k-1}{k}$.
\end{thm}

\begin{proof}
  Let $(Q_x: x\in V(I))$ denote the given $I$-partition of $G$ and let $\langle V_1,\ldots,V_h\rangle$ be the given layering of $G$.  For a node $x\in V(T)$, let $R_x:=\bigcup_{y\in C_x} Q_{y}$.
  Apply \lemref{nice-decomposition} to $I$ to obtain a tree-decomposition $(C_x: x\in V(T))$ of $I$ where $T$ is rooted at some node $x_0$ and that satisifies Properties~\tref{t:first}--\tref{t:last}. (We skip the introduction of the isomorphic graph $H'$ and assume from the outset that $V(I)\subseteq V(T))$.)

  For each $v\in V(G)$, let $q(v)$ denote the unique node $q\in V(I)$ such that $v\in Q_q$ and let $c(v)$ denote the root of the subtree $T[v]:=T[\{x\in V(T):q(v)\in C_x]$. For each $x\in V(T)$, define the following objects:
  \begin{enumerate}
    \item $Y_x=\{v\in V(G):c(v)=x\}$.  
    \item $V_x=\bigcup_{y\in V(T_x)} V(Y_y)$.  
    \item $F_x = \{w\in V(G): vw\in E(G), w\in V_x$. 
    \item $N_x = G[V_x\cup V(F_x)]$.  
  \end{enumerate}

  These objects have the following properties:

  \begin{enumerate}
    \item $(Y_x: x\in V(T))$ is a partition of $V(G)$.
    \item For all $i\in\{1,\ldots,h\}$, $|Y_x\cap L_i| \le t\ell$. (Since $Y_x\subseteq\bigcup_{y\in C_x} Q_y$, $|C_x|\le t$, and $|Q_y\cap L_i|\le\ell$.)
    \item $N_x\subseteq N_{y}$ for every $T$-ancestor $y$ of $x$.
    \item $x$ has $t'\le t$ $T$-ancestors $a_1,\ldots,a_{t'}$ such that $F_x \subseteq \bigcup_{i=1}^{t'} Y_{a_i}$.
  \end{enumerate}
  Property~4 follows from \tref{t:ancestor-edge}:  For each edge $vw\in E(G)$ with $w\in V_x$ and $v\in Y_{a_i}$, the graph $I$ contains the edge $q(v)q(w)=a_ix'$ where (by Property~3) $x'$ is a $T$-descendant of $x$ (possibly $x'=x$).  By \tref{t:ancestor-edge} one of $x'$ or $a_i$ is a $T$-ancestor of the other. For some reason (????) $a_i$ is an ancestor of $x'$.  Therefore $a_i$ is contained in $C_x$.  Property~4 then follows from the fact that $|C_x|\le t$.

  =====

  Let $(Q_x: x\in V(I))$ denote the given $I$-partition of $G$.  For a node $x\in V(T)$, let $R_x:=\bigcup_{y\in C_x} Q_{y}$. For a node $x\in V(T)$, we say that a $(k,d)$ $vw$-shortcut $P$ in $G$ \emph{touches} $x$ if $V(P)\cap R_x\neq\emptyset$.  If $G$ contains a $vw$-shortcut that touches $x$, then we say that $v$ and $w$ \emph{participate} in $x$.

  For any $v\in V(G)$, the union of all $(k,d)$ shortcuts in $G$ that have $v$ as an endpoint is a connected subgraph of $G$.  It follows that the set of nodes $x\in V(T)$ in which $v$ participates is a non-empty (connected) subtree of $T$ and we let $a(v)$ denote the root of this subtree. For each $x\in V(T)$, we define the \emph{separator}
  \[  
       S_x := \{v\in V(G): a(v)= x\} \enspace .
  \]
  Observe that $(S_x : x\in V(T))$ is a partition of $V(G)$.  We let $H:=G^{k,d}/(S_x : x\in V(T))$ denote the resulting quotient graph and we let $V(H)=V(T)$ in the obvious way, so $x\in V(H)$ is the vertex obtained by contracting $S_x$ in $G^{k,d}$.

  For a node $x$ in $T$ with ancestors $x_0,x_1,\ldots,x_r=x$ that roots a maximal subtree $T_x$, we define several subgraphs of $G$:  $\overline{G}_x = G[\bigcup_{y\in V(T_x)} R_y$, $G_x=\overline{G}_x\setminus(\bigcup_{i=1}^{r-1} R_{x_i})$, $F_x=G[\{w\in V(G): vw\in E(G), v\in V(G_x)\}]$, $Y_x=G[\{v\in V(G): a(v)=x]$

   $F_x=G[R_x\setminus \bigcup_{i=1}^{r-1} R_{x_i}]$.  We let $T_x$ denote the subtree of $T$ rooted at $x$ and let $G_x := G[\bigcup_{y\in V(T_x)} R_y\setminus R_x]$.

  We claim that, for any edge $xy\in E(H)$, $x$ is a $T$-ancestor of $y$ or vice-versa.  Indeed, if $xy\in E(H)$, then there exists an edge $vw\in E(G^{k,d})$ with $v\in S_x$ and $w\in S_y$.

  Therefore $G$ contains a $vw$-shortcut $P$ where $v\in Q_{x'}$ for some $T$-descendant $x'$ of $x$ and $w\in Q_{y'}$ for some $T$-descendant $y'$ of $y$.  If neither $x$ nor $y$ is were a $T$-ancestor of the other, then neither $x'$ nor $y'$ would be an ancestor of the the other, but both $v$ and $w$ would participate in a node $z$ that is a $T$-ancestor of both $x$ and $y$.  This would contradict the fact that $v\in S_x$ and $w\in S_y$.

  We construct a tree-decomposition $\mathcal{T}:=(B_x:x\in V(T))$ of $I$ as follows: For each edge $xy\in E(I)$ we include $x$ in $B_z$ for each node $z$ on the path, in $T$ from $x$ to $y$.  This is clearly a tree-decomposition of $I$ and it has the property that, for any $x\in V(I)$, $T[x]$ is rooted at $x$.  All that remains is to bound the width of $\mathcal{T}$.

  We now focus on one particular bag $B_x$ for some $x\in V(T)$.  
  Let $x_0,\ldots,x_d$ be the path from $x_0:=x$ to the root of $T$.  
  Let $T_i$ denote the subtree of $T$ rooted at $x_i$ and let
  \[ 
      G_i := G[\bigcup_{x'\in V(T_i)}\bigcup Q_{x'}]-Q_{x_{i-1}} \enspace .
  \]
  In other words, $G_i$ is the part of $G$ that contains that is separated by the removal of $S$

  Define $N_{x_0}:=G$ and, for each $i\in\{1,\ldots,d\}$, define $N_{x_i}$ to be the subgraph of $G$ induced by the union, over all $x'$ in the subtree of

  % containing exactly those components of $N_{x_{i-1}}-Q_{x_i}$ that contain vertices in $

  As discussed above, for any node $x\in V(T)$, the bag $B_x$ contains only $T$-ancestors of $x$ (including $x$ itself).  If $B_x$ contains some $T$-ancestor $y$ then this is because $xy\in E(I)$ which implies that there exists $vw\in E(G')$ with $v\in S_{x'}$ and $w\in S_y$ for some $T$-descendant $x'$ of $x$.  This implies that $G$ contains a $vw$-shortcut $P=v_0,\ldots,v_p$ with $v_0=v$ and $v_p=w$.   
  Note that neither $v$ nor $w$ participates any $T$-ancestor of $y$ other than $y$.  Therefore $P$ does not contain any vertices of $Q_{y'}$ for any ancestor $y'$ of $y$.
\end{proof}
